Default GMSE data structures

### 2020-05-29 notebook/cran_vignettes/SI6.Rmd

---
title: "Management frequency and extinction risk"
author: A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;,
  Jeroen Minderman&#xb9;, Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy
  Rakotonarivo&#xb9;, Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting
  Information 6)'
biblio-style: apalike
---

@@

The default sub-models of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which in GMSE are represented by individual table rows (as in `RESOURCES`, `AGENTS`, and `OBSERVATION`) or layers of three-dimensional arrays (as in `COST` and `ACTION`). Individual-based models (IBMs) have been a useful approach in ecology for decades [@Uchmanski1996; @Grimm1999], providing both a pragmatic tool for the mechanistic modelling of complex populations and a powerful technique for theoretical investigation. A key advantage of the individual-based modelling approach is the discrete nature of individuals, which allows for detailed trait variation and complex interactions among individuals. In GMSE, some of the most important traits for resources include types, ages, demographic parameter values, locations, etc., and for agents (manager and users), traits include different types, utilities, budgets, etc. The traits that resources and managers have can potentially affect their interactions, and default GMSE sub-models take advantage of this by simulating interactions explicitly on a landscape (see [SI7](SI7.pdf) for an introduction to GMSE default data structures).

>>>>

---
title: "Management frequency and extinction risk"
author: |
        | A. Bradley Duthie&#xb9;&#xb3;, Jeremy J. Cusack&#xb9;, Isabel L. Jones&#xb9;, Jeroen Minderman&#xb9;, 
        | Erlend B. Nilsen&#xb2;, Roc&#0237;o A. Pozo&#xb9;, O. Sarobidy Rakotonarivo&#xb9;, 
        | Bram Van Moorter&#xb2;, and Nils Bunnefeld&#xb9;
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling, UK [2] Norwegian Institute for Nature Research, Trondheim, Norway [3] alexander.duthie@stir.ac.uk'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document:
    fig_caption: yes
    reference_docx: docx_template.docx
header-includes:
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: gmse.bib
subtitle: 'GMSE: an R package for generalised management strategy evaluation (Supporting
  Information 6)'
biblio-style: apalike
---

@@

The default sub-models of GMSE (`resource`, `observation`, `manager`, `user`) are individual-based (also called 'agent-based'), meaning that they model discrete individuals (resources or agents), which in GMSE are represented by individual table rows (as in `RESOURCES`, `AGENTS`, and `OBSERVATION`) or layers of three-dimensional arrays (as in `COST` and `ACTION`). Individual-based models (IBMs) have been a useful approach in ecology for decades [@Uchmanski1996; @Grimm1999], providing both a pragmatic tool for the mechanistic modelling of complex populations and a powerful technique for theoretical investigation. A key advantage of the individual-based modelling approach is the discrete nature of individuals, which allows for detailed trait variation and complex interactions among individuals. In GMSE, some of the most important traits for resources include types, ages, demographic parameter values, locations, etc., and for agents (manager and users), traits include different types, utilities, budgets, etc. The traits that resources and managers have can potentially affect their interactions, and default GMSE sub-models take advantage of this by simulating interactions explicitly on a landscape (see [Default GMSE data structures](SI7.pdf) for an introduction to GMSE default data structures).

### 2020-05-29 notebook/website_vignettes/SI3.Rmd

To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](https://www.biorxiv.org/content/biorxiv/early/2017/11/17/221432.full.pdf#page=5) in the main text. We consider a protected population of waterfowl that exploits and damages agricultural land and is therefore a source of conservation conflict between those that seek the conservation of waterfowl and those that are concerned with the loss of agriculture [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018]. As in the main text example, the objective of the manager is to keep waterfowl at a target abundance that minimises extinction risks, while the objective of farmers is to maximise agricultural production on their landscape. Here we consider a more complex simulation, with a level of detail that more accurately reflects a scenario that might occur in a real social-ecological system where the manager sets policies that incentivise the user to act in a way that ensures the persistence of the resource. The policies are backed up by a manager budget that can be allocated to set costs of actions and thus incentivise users to perform the desired actions. The user also has a budget to carry out actions, and the cost of the user actions is affected by the manager policies and budget. Our objective is not to model the dynamics of a specific system, but to show how GMSE could be parameterised using demographic estimates from empirical studies. We therefore consider an example population in which such estimates are well-reported and readily available.

@@

Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible in GMSE. For simplicity, here we assess conflict using only the `gmse` function to show how parameter values can be set to provide useful results. Novice R users may prefer to run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line. Alternatively, experienced R users may prefer to simulate by [looping time steps](SI2.pdf#page=5) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values.

Simulations using the default GMSE sub-models described above are run using the `gmse` function, which offers a range of options for setting parameter values (see Table 1 for some select examples). Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation (see [SI7](SI7.pdf) for a description of key data structures).

@@

Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016. Where GMSE parameter values are not available, we use reasonable values or GMSE defaults. To make model inferences for real case studies, we strongly recommend [replicating simulations](SI6.pdf) and simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.

@@

The code below calls `gmse` using the empirically derived parameters for Taiga Bean Geese described above. We also set `manager_budget = 10000` and `user_budget = 10000`. Further, we consider the case of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be interpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`). Landscape size is set to default 100 by 100 cells, so each farmer owns about 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the [advanced GMSE options](SI4.pdf) using `gmse_apply`). Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`. We assume that a single goose decreases agricultural production on a cell by 2% per time step (`res_consume = 0.02`). We further assume that the population is very well-monitored, with observers counting goose numbers on each cell of the landscape in every timestep (`observe_type = 3`) with the ability to observe one landscape cell in every direction (`agent_view = 1`). All other parameter values are set to GMSE defaults.

@@

The GMSE function `gmse` and its graphical user interface counterpart `gmse_gui` offer wide a suite of options for parameterising simulations to fit empirical case studies of conservation interest using default GMSE natural resource, observation, manager, and user sub-models. Future development of these sub-models might usefully incorporate additional details relevant to specific case studies, such population structure, multiple natural resource and user types, or different manager policy and user action possibilities. [Requests for new features](https://github.com/bradduthie/gmse/issues) and contributions to GMSE can be made through [GitHub](https://github.com/bradduthie/gmse). Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](https://bradduthie.github.io/gmse_SI2.pdf) to more flexibly simulate social-ecological systems. In [SI4](SI4.pdf), we show an example of this using the same Taiga Bean Geese case study that we did here.

>>>>

To demonstrate how GMSE can be used to understand conflict in social-ecological systems, we build upon the [example of resource management](GMSE.html#an-example-of-resource-management) in the main text. We consider a protected population of waterfowl that exploits and damages agricultural land and is therefore a source of conservation conflict between those that seek the conservation of waterfowl and those that are concerned with the loss of agriculture [e.g., @Fox2017a; @Mason2017; @Tulloch2017; @Cusack2018]. As in the main text example, the objective of the manager is to keep waterfowl at a target abundance that minimises extinction risks, while the objective of farmers is to maximise agricultural production on their landscape. Here we consider a more complex simulation, with a level of detail that more accurately reflects a scenario that might occur in a real social-ecological system where the manager sets policies that incentivise the user to act in a way that ensures the persistence of the resource. The policies are backed up by a manager budget that can be allocated to set costs of actions and thus incentivise users to perform the desired actions. The user also has a budget to carry out actions, and the cost of the user actions is affected by the manager policies and budget. Our objective is not to model the dynamics of a specific system, but to show how GMSE could be parameterised using demographic estimates from empirical studies. We therefore consider an example population in which such estimates are well-reported and readily available.

@@

Our goal is not to provide a detailed case study of the Taiga Bean Geese, but rather to demonstrate how such a case study would be possible in GMSE. For simplicity, here we assess conflict using only the `gmse` function to show how parameter values can be set to provide useful results. Novice R users may prefer to run all of the simulations below using the browser-based GMSE GUI by calling `gmse_gui()` from the R command line. Alternatively, experienced R users may prefer to simulate by [looping time steps](SI2.html#running-gmse-simulations-by-looping-gmse_apply) through `gmse_apply`, which allows more flexibility for incorporating custom sub-models and dynamically adjusting parameter values.

Simulations using the default GMSE sub-models described above are run using the `gmse` function, which offers a range of options for setting parameter values (see Table 1 for some select examples). Output of `gmse` is an exhaustive list that includes all resources and observations, all stakeholder decisions and actions, and all landscape properties in each time step of the simulation (see [Default GMSE data structures](SI7.html) for a description of key data structures).

@@

Where available, we use estimated demographic parameter values from @Johnson2018 and @AEWA2016. Where GMSE parameter values are not available, we use reasonable values or GMSE defaults. To make model inferences for real case studies, we strongly recommend [replicating simulations](SI6.html) and simulating across a range of parameter values when empirical estimates are unavailable, as social-ecological dynamics might be sensitive to these unknown values.

@@

The code below calls `gmse` using the empirically derived parameters for Taiga Bean Geese described above. We also set `manager_budget = 10000` and `user_budget = 10000`. Further, we consider the case of a region in which farmland makes up 60% of all land, with 40% of land being 'public' (`public_land = 0.4`; which might be interpreted as any land in which stakeholders are not, or cannot be, invested in goose presence), and divide the farmland amongst 80 individual farmers (`stakeholders = 80`; `land_ownership = TRUE`). Landscape size is set to default 100 by 100 cells, so each farmer owns about 75 cells, which might be interpreted as hectares of land (for instructions on how to more precisely control landscape ownership, see the [advanced GMSE options](SI4.html) using `gmse_apply`). Because we need both density-dependent (`res_death_K = 93870`) and density-independent (`remove_pr = 0.122`) sources of mortality, we set `res_death_type = 3`. We assume that a single goose decreases agricultural production on a cell by 2% per time step (`res_consume = 0.02`). We further assume that the population is very well-monitored, with observers counting goose numbers on each cell of the landscape in every timestep (`observe_type = 3`) with the ability to observe one landscape cell in every direction (`agent_view = 1`). All other parameter values are set to GMSE defaults.

@@

The GMSE function `gmse` and its graphical user interface counterpart `gmse_gui` offer wide a suite of options for parameterising simulations to fit empirical case studies of conservation interest using default GMSE natural resource, observation, manager, and user sub-models. Future development of these sub-models might usefully incorporate additional details relevant to specific case studies, such population structure, multiple natural resource and user types, or different manager policy and user action possibilities. [Requests for new features](https://github.com/bradduthie/gmse/issues) and contributions to GMSE can be made through [GitHub](https://github.com/bradduthie/gmse). Additionally, where entirely different types of sub-models are required, the `gmse_apply` function [can be used](SI2.html) to more flexibly simulate social-ecological systems. In [Advanced case study options](SI4.html), we show an example of this using the same Taiga Bean Geese case study that we did here.

References
================================================================================

### 2020-05-29 notebook/website_vignettes/SI7.Rmd

The default sub-models of GMSE (`resource`, `observation`, `manager`, and `user`) use a small number of default data structures to hold the information needed in simulations. While these default sub-models do not necessarily need to be used in every run in GMSE (see [use of](SI2.pdf) `gmse_apply`), they will be used in any run of the `gmse` function, and in any call of the `gmse_apply` function that does not run with entirely custom sub-models. Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules, as in [SI4](SI4.pdf)), or build custom individual-based sub-models. Here we provide a brief explanation of the following key data structures (each name below is listed as it is named in the output `gmse_apply` when `get_res = "Full"`).

@@

Note that these are not the only data structures used in GMSE, but they are the only ones that can be easily modified in GMSE v0.4.0.7 (see, e.g., [SI4](SI4.pdf)), so they are the ones that we focus on here. Additionally, any custom subfunction that returns an array rather than a single value should adhere to the same structure as these defaults if any default GMSE functions are to be used in `gmse_apply`. We can investigate each data structure by running a single simulation of `gmse_apply`.

@@

In the [genetic algorithm](SI1.pdf), values in elements of a `user_array` layer are potentially modified according to each agent's objective, as constrained by costs in `manager_array`.

@@

In [SI4](SI4.pdf), we show how landscape cell values can be manipulated to customise the placement of land ownership. By default, GMSE v0.6.0.0+ places the ownership of landscape cells using a shortest-splitline algorithm (see the lab notebook for [a detailed explanation](https://confoobio.github.io/gmse/notebook/gmse_notes.html#u-10-MAY-2020)). The algorithm splits the landscape evenly into smaller and smaller rectangles until a sufficient number of rectangles are produced for all users to own a similar amount of land (note that how uniformly land is partitioned among users will naturally be influenced by how evenly the total number of landscape cells can be divided among them). Below is an example in which land of `land_dim_1 = 64` and `land_dim_2 = 64` is partitioned equally among eight users.

@@

In addition, [whereas in previous versions variable landscape ownership among users had to be customised using `gmse_apply`](SI4.pdf), GMSE v. 0.6.0.0+ now includes an additional parameter `ownership_var`, which takes effect when `land_ownership = TRUE`. The value of this parameter is expected to be >=0 and <1. At its default value of 0, land ownership is equal among users as demonstrated in the code above. Values >0 increase the extent of variability of landownership among users; in technical terms, `ownership_var` controls the extent to which each split in the shortest-splitline algorithm is proportional to the number of users the landscape is to be divided into, with larger numbers shifting the split away from even distribution.
In practical terms, this means that increasing values of `ownership_var` produce more unevenly distributed landscapes among users. For example, a value of `ownership_var = 0.1` produces a relatively limited about of inequity in landscape distribution among users.

@@

It should be noted that it is still entirely possible for the user to customise landscapes to whatever extent necessary, using custom `gmse_apply` loops [as described previously](SI4.pdf); however, the new `ownership_var` provides an option in GMSE to alter landscape ownership in a relatively straightforward way.

@@

We have focused on the data structures [`AGENTS`](#agents), [`resource_array`](#resources), [`observation_array`](#observation), [`manager_array`](#cost), [`user_array`](#action), and [`LAND`](#land) because these are the data structures that can be most readily manipulated to customise GMSE simulations. An example of how to do this within a loop using `gmse_apply` can be found in [SI4](SI4.pdf). While other data structures exist within GMSE (e.g., see the output of `gmse_apply` when `get_res = "Full"`), we do not recommend manipulating these structures for custom simulations.

Many data structures contain elements that are unused in GMSE v0.6.0.0, and in all cases this is designed for ease of ongoing development of new GMSE features. Requests for new features can be made on GitHub using the [GMSE Wiki](https://github.com/bradduthie/gmse/wiki) or the [GMSE Issues](https://github.com/bradduthie/gmse/issues) page.

>>>>

The default sub-models of GMSE (`resource`, `observation`, `manager`, and `user`) use a small number of default data structures to hold the information needed in simulations. While these default sub-models do not necessarily need to be used in every run in GMSE (see [use of](SI2.html) `gmse_apply`), they will be used in any run of the `gmse` function, and in any call of the `gmse_apply` function that does not run with entirely custom sub-models. Simulation and model inference do not require an understanding of the default data structures, but such an understanding can be especially useful when running `gmse_apply` if there is a need to extract uncommonly used information, change key simulated values (e.g., landscape properties, agent budgets, or resource movement rules, as in [Advanced case study options](SI4.html)), or build custom individual-based sub-models. Here we provide a brief explanation of the following key data structures (each name below is listed as it is named in the output `gmse_apply` when `get_res = "Full"`).

@@

Note that these are not the only data structures used in GMSE, but they are the only ones that can be easily modified in GMSE (see, e.g., [Advanced case study options](SI4.html)), so they are the ones that we focus on here. Additionally, any custom subfunction that returns an array rather than a single value should adhere to the same structure as these defaults if any default GMSE functions are to be used in `gmse_apply`. We can investigate each data structure by running a single simulation of `gmse_apply`.

@@

In the [genetic algorithm](SI1.html), values in elements of a `user_array` layer are potentially modified according to each agent's objective, as constrained by costs in `manager_array`.

@@

In [Advanced case study options](SI4.html), we show how landscape cell values can be manipulated to customise the placement of land ownership. By default, GMSE v0.6+ places the ownership of landscape cells using a shortest-splitline algorithm (see the lab notebook for [a detailed explanation](https://confoobio.github.io/gmse/notebook/gmse_notes.html#u-10-MAY-2020)). The algorithm splits the landscape evenly into smaller and smaller rectangles until a sufficient number of rectangles are produced for all users to own a similar amount of land (note that how uniformly land is partitioned among users will naturally be influenced by how evenly the total number of landscape cells can be divided among them). Below is an example in which land of `land_dim_1 = 64` and `land_dim_2 = 64` is partitioned equally among eight users.

@@

In addition, [whereas in previous versions variable landscape ownership among users had to be customised using `gmse_apply`](SI4.html), GMSE v. 0.6+ now includes an additional parameter `ownership_var`, which takes effect when `land_ownership = TRUE`. The value of this parameter is expected to be >=0 and <1. At its default value of 0, land ownership is equal among users as demonstrated in the code above. Values >0 increase the extent of variability of landownership among users; in technical terms, `ownership_var` controls the extent to which each split in the shortest-splitline algorithm is proportional to the number of users the landscape is to be divided into, with larger numbers shifting the split away from even distribution.
In practical terms, this means that increasing values of `ownership_var` produce more unevenly distributed landscapes among users. For example, a value of `ownership_var = 0.1` produces a relatively limited about of inequity in landscape distribution among users.

@@

It should be noted that it is still entirely possible for the user to customise landscapes to whatever extent necessary, using custom `gmse_apply` loops [as described previously](SI4.html); however, the new `ownership_var` provides an option in GMSE to alter landscape ownership in a relatively straightforward way.

@@

We have focused on the data structures [`AGENTS`](#agents), [`resource_array`](#resources), [`observation_array`](#observation), [`manager_array`](#cost), [`user_array`](#action), and [`LAND`](#land) because these are the data structures that can be most readily manipulated to customise GMSE simulations. An example of how to do this within a loop using `gmse_apply` can be found in [Advanced case study options](SI4.html). While other data structures exist within GMSE (e.g., see the output of `gmse_apply` when `get_res = "Full"`), we do not recommend manipulating these structures for custom simulations.

Many data structures contain elements that are unused in GMSE v0.6, and in all cases this is designed for ease of ongoing development of new GMSE features. Requests for new features can be made on GitHub using the [GMSE Wiki](https://github.com/bradduthie/gmse/wiki) or the [GMSE Issues](https://github.com/bradduthie/gmse/issues) page.
